---
title: "Assignment Part 2: new ML model"
author: "Quinn Thomas"
date: "`r Sys.time()`"
output: github_document
---

```{r message=FALSE}
library(tidyverse)
library(tidymodels)
tidymodels_prefer()

#For model
library(caret)
library(xgboost)
library(parsnip)
library(recipes)

```

## Objective

Using machine learning and tidymodels to predict how much carbon is stored in vegetation across the U.S.

## Expections

This assignment requires the following

- You must use more predictors than used in the `machine-learning-101.Rmd` and `example-with-tuning.Rmd`.  You do not have to use all predictors.  In fact, one of the decisions that you will be making is which predictors you will include.
  
- You must use a different model and/or engine than used in the `machine-learning-101.Rmd` and `example-with-tuning.Rmd`.

- You must upload your predictions of the new carbon stock data to Canvas as a csv.
  
### Guidance

You will need to explore the different models/engines that are avialable in the tidymodels package.  Following the the link below will show the list of models that are avialable.  Clicking on a model type will show the different engines that are avialable to use.  

**[Alternative modeling approaches](https://parsnip.tidymodels.org/reference/index.html#models)**

Here is a list of models that may be particularly useful for the data in this assignment.

- [Single layer neural network](https://parsnip.tidymodels.org/reference/mlp.html)
- [Random forest](https://parsnip.tidymodels.org/reference/rand_forest.html)
- [Automatic Machine Learning](https://parsnip.tidymodels.org/reference/auto_ml.html)
- [Boosted trees](https://parsnip.tidymodels.org/reference/boost_tree.html)
- [Linear Regression](https://parsnip.tidymodels.org/reference/linear_reg.html)

The documentation of the engine will describe any pre-processessing that is required for the data and list the hyperparameters that are avialable for tuning. For example [here](https://parsnip.tidymodels.org/reference/details_rand_forest_ranger.html) is the documentation for the `ranger` engine within the `rand_forest` model

Once you start pre-processing your data using a recipe, the following link will show the different "steps" that you can use.  Clicking on the step will show how to use it in a recipe.  

**[Different recipes steps for feature engineering](https://recipes.tidymodels.org/reference/index.html)**

There are broad classes of recipe steps that you might find useful for the data that you are working with in this assignment.  You will be not be using all the available steps.

- [Imputation (fill NA values)](https://recipes.tidymodels.org/reference/index.html#step-functions-imputation)
- [Transform variables](https://recipes.tidymodels.org/reference/index.html#step-functions-individual-transformations)
- [Add dummy variables](https://recipes.tidymodels.org/reference/index.html#step-functions-dummy-variables-and-encodings)
- [Create variable interaction](https://recipes.tidymodels.org/reference/index.html#step-functions-interactions)
- [Normalize variables (e.g., center on 0 and scale sd)](https://recipes.tidymodels.org/reference/index.html#step-functions-normalization)
- [Filter rows and select columns](https://recipes.tidymodels.org/reference/index.html#step-functions-filters)
- [Operate on rows](https://recipes.tidymodels.org/reference/index.html#step-functions-row-operations)

## Step 0: Warm-up

**Question 11:** What model and engine did you select to use?

**Answer 11: The model I chose to use was the boost_tree() model with the xgboost engine.**

**Question 12:** Provide a high level description of the modeling approach

**Answer 12: To begin my modeling approach, I first read-in the biomass data and added a column called "ID" that assigned a numerical label to each nlcdClass since the xgboost is not able to translate factor predictors, so all predictors must be converted to numeric values. I then split the data into training and testing data sets with an 80:20 ratio, respectfully. I also used the set.seed() function to ensure that the different attempts/run of code had the same training data to ensure that the code I altered was affecting the efficiency of the model, rather than the random data set generated. I then added folds to the training data to further train my model. I then made the recipe for my model and removed the "plotID" and "nlcdClass" columns in step_rm from the model because they are not numeric columns. I also added step_impute_mean() and step_normalize() during trial and error code, and kept those as they decreased the rmse. Next I defined my model and added the tuning parameters tree_depth = and trees = which affect the amount and detail of trees created for this model. I also set the mode as "regression" since the goal is to predict data for a separate, but similar dataset. Next the workflow was assembled using the workflow() function and the previously created recipe and workflow. Next, the hyperparameters were tuned, specifically running the training code through the folds a described amount of times, grid function, and selecting "rmse" as a desired metric. The workflow was then updated with the new tuned hyper-parameters by selecting the best parameter values that provided the rmse value. The model was then fit to the training data which could then be used to predict values for the test data. Finally, the model was used to evaluate the new dataset and predictions were made for those plotIDs. **


**Question 13:** Does the model require any special pre-processing of data. If so describe the steps needed.

**Answer 13: The model required all predictors to be in numeric form, so in order to use the nlcdClass predictor, I created numerical labels for each unique nlcdClass and used that column when training my model. Other than having all predictors be numerical, there are no other necessary pre-processing steps needed in order for this model to work at the basic level. **

**Question 14:** List any hyper-parameters that can be tuned

**Answer 14: Listed below are all the hyper-parameters that can be tuned listed on the parsnip tidymodels site for the xgboost engine. For my model, I used trees =  and tree_depth =. I tuned other combinations of hyper-parameters, but these gave the best results.**
#   mtry = integer()
#   trees = integer()
#   min_n = integer()
#   tree_depth = integer()
#   learn_rate = numeric()
#   loss_reduction = numeric()
#   sample_size = numeric()
#   stop_iter = integer()

## Step 1: Obtain data

```{r}
biomass_data_original <- read_csv("data/neon_biomass.csv", show_col_types = FALSE)
```

```
#original 
biomass_data <- biomass_data_original |> 
  select(plotID, nlcdClass, plot_kgCm2, precip, tavg) |> 
  mutate(ID = as.numeric(as.factor(nlcdClass)))
```

```{r}
#submitted
biomass_data1 <- biomass_data_original |> 
  select(plotID, nlcdClass, plot_kgCm2, precip, vpd) |> 
  mutate(ID = as.numeric(as.factor(nlcdClass)))

```

```
spc_tbl_ [1,000 × 17] (S3: spec_tbl_df/tbl_df/tbl/data.frame)
 $ plotID    : chr - identifier of plot
 $ daylength : num - mean annual daylength (s/day)
 $ precip    : num - mean annual precipitation (mm)
 $ range     : num - mean annual daily range in temperature (day max - day min; C)
 $ tavg      : num - mean annual temperature (C)
 $ solar     : num - mean annual solar radiation (W/m2)
 $ tmax      : num - mean annual daily max temperature (C)
 $ tmin      : num - mean annual daily min temperature (C)
 $ vpd       : num - mean annual vapor pressure deficient (Pa)
 $ elevation : num - elevation of plot (meters)
 $ nlcdClass : chr - land-cover classification of plot
 $ lat       : num - latitude of plot (degrees W)
 $ long      : num - longitude of plot (degrees E)
 $ plotType  : chr - type of plot (tower or distributed)
 $ ndvi      : num - normalized difference vegetation index of plot from MODIS (unitless)
 $ siteID    : chr - identifier of site that includes the plot
 $ plot_kgCm2: num - vegetation carbon stock of plot (kgC/m2)
```

## Step 2: Pre-process data

### Split data into training/testing sets

**Question 15:** Provide code for splitting data

**Answer 15:**

```
#original
set.seed(100)
split <- initial_split(biomass_data, prop = 0.80, strata = ID)

train_data <- training(split)
test_data <- testing(split)
```

```{r}
#submitted
set.seed(100)
split1 <- initial_split(biomass_data1, prop = 0.80, strata = ID)

train_data1 <- training(split1)
test_data1 <- testing(split1)
```

### Split training data into folds

This step is only required if you are using a model and engine that has parameters that can be tuned.

**Question 16:** Provide code for splitting data. If your model does not require parameter tuning, then state that as your answer rather than providing the code.

**Answer 16:**

```
#original
folds <- vfold_cv(train_data, v = 10)
```

```{r}
#submitted
folds1 <- vfold_cv(train_data1, v = 10)

```

### Feature engineering using a recipe

**Question 17:** Provide code that defines the recipe for feature engineering. Be sure to follow the recommendations for your selected engine. 

**Answer 17:**

```
#original
biomass_recipe <- biomass_data |> 
  recipe(plot_kgCm2 ~ . ) |> 
  step_rm(plotID, nlcdClass) |>
  step_impute_mean() 
```
```{r}
#submitted
biomass_recipe1 <- biomass_data1 |> 
  recipe(plot_kgCm2 ~ . ) |> 
  step_rm(plotID, nlcdClass) |> 
  step_impute_mean() |> 
  step_normalize()

```


## Step 3: Specify model and workflow

### Define model type and mode

**Question 18:** Provide code that defines your model (model type + engine).

**Answer 18:**

```
### TUNING ###
#   mtry = integer()
#   trees = integer()
#   min_n = integer()
#   tree_depth = integer()
#   learn_rate = numeric()
#   loss_reduction = numeric()
#   sample_size = numeric()
#   stop_iter = integer()

#original
boost_for_mod <- 
  boost_tree() |>
  set_engine("xgboost") |> 
  set_mode("regression")
```

```{r}
#submitted
boost_for_mod1 <- 
  boost_tree(
    tree_depth = tune(),
    trees = 1500) |>
  set_engine("xgboost") |> 
  set_mode("regression")
```


### Define workflow

**Question 19:** Provide code that defines the workflow

**Answer 19:**

```
#original
biomass_wflow <-
  workflow() |> 
  add_model(boost_for_mod) |> 
  add_recipe(biomass_recipe)

biomass_wflow
```

```{r}
#submitted
biomass_wflow1 <-
  workflow() |> 
  add_model(boost_for_mod1) |> 
  add_recipe(biomass_recipe1)

biomass_wflow1


```

## Step 4: Train model on Training Data

### Estimate best hyper-parameters using tuning

**Question 20:** Provide code that tunes the hyper-parameters of your model.  If your model+engine does not require tuning, then state that as your answer.

**Answer 20:**

```{r}
#submitted
biomass_resample_fit1 <- 
  biomass_wflow1 |>  
  tune_grid(resamples = folds1,
            grid = 25,
            control = control_grid(save_pred = TRUE),
            metrics = metric_set(rmse))

```

### Update workflow with best hyper-parameters

**Question 21:** Provide code that updates the workflow with the best hyper-parameters. If your model+engine does not require tuning, then state that as your answer.

**Answer 21:**

```{r}
#submitted
biomass_resample_fit1 %>% 
  collect_metrics() |> 
  arrange(mean)

best_hyperparameters1 <- biomass_resample_fit1 %>%
  select_best("rmse")

best_hyperparameters1

final_workflow1 <- 
  biomass_wflow1 %>% 
  finalize_workflow(best_hyperparameters1)


```

## Step 5: Fit to all training data

**Question 22:** Provide code that fits your model to the training data. Use the best hyper-parameter if you tuned them.

**Answer 22:**

```
#original
biomass_fit <- 
  biomass_wflow |> 
  fit(data = train_data)
```
```{r}
#submitted
biomass_fit1 <- 
  final_workflow1 |> 
  fit(data = train_data1)

```

## Step 6: Predict Test Data

**Question 23:** Provide code that predicts the test data

**Answer 23:**

```
#original
predictions <- predict(biomass_fit, new_data = test_data)

pred_test <- bind_cols(test_data, predictions)
```
```{r}
#submitted
predictions1 <- predict(biomass_fit1, test_data1)

pred_test1 <- bind_cols(test_data1, predictions1)

```

## Step 7: Evaluate model

**Question 24:** Provide code calculates the rmse and r2 for the test data.  Make sure that the table with rmse and r2 is shown in your knitted document.

**Answer 24:**

```
#original ## 
multi_metric <- metric_set(rmse, rsq)

metric_table <- pred_test |> 
multi_metric(truth = plot_kgCm2, estimate = .pred)

metric_table
```

```{r}
#submitted ## 
multi_metric1 <- metric_set(rmse, rsq)

metric_table1 <- pred_test1 |> 
multi_metric1(truth = plot_kgCm2, estimate = .pred)

metric_table1


```

## Step 8: Deploy model

Use your trained model to predict new data. The new data is a data frame with all the same columns as the training and testing set above but with `NA` values for the biomass. I have the true biomass values and will calculate rmse for your predictions. Therefore you need to predict the biomass and upload your csv to Canvas and I will download them as a set and run the analysis in class to examine the "scores"

Be sure to change the `name` to be your last name.

```{r}
name <- "Chiang"
```

```{r}
submit_data <- read_csv("data/neon_biomass_new.csv", show_col_types = FALSE) |> 
  mutate(ID = as.numeric(as.factor(nlcdClass)))

```


**Question 25:** Provide code to predict the new data

**Answer 25:**
```{r}
submit_predictions <- predict(biomass_fit1, new_data = submit_data) 

submit_predicted <- bind_cols(submit_data, submit_predictions) |>
  select(plotID, nlcdClass, .pred, precip, vpd)

```

**Question 26:** Use `write_csv(submit_predicted, file = paste0(name,"-submission.csv"))` to write your submissions to a csv.

**Answer 26:**
```{r}
write_csv(submit_predicted, file = paste0(name,"-submission.csv"))

```


**Question 27:** Upload `neon_biomass_submission.csv` to Canvas.

**Answer 27:**

## Knitting and committing

Remember to Knit your document as a `github_document` and comment+push to GitHub your code, knitted document, and any files in the `figure-gfm` subdirectory that was created when you knitted the document.

